Synthesis of Linear Functional Programs

### 2021-05-09 article/article.tex
init goals and work plan

\section{Separate}

More about the concepts than citing concrete works (but citations are
needed!):
\begin{enumerate}
\item PL/Type systems in a ``formal'' sense (inference rules, etc) [DONE]
\item Propositions as types: Proofs as programs, proof search as
  synthesis, etc. [MOSTLY DONE]
\item Linear logic: a ``resource aware logic'' [ONGOING]
\item Proof search in LL: challenges, focusing as a ``solution'' [ONGOING]
\item Linear $\lambda$-calculus: what does it look like.
\item Type-based synthesis: ``inverting'' a type system.
\end{enumerate}

@@

Refinement types provide very precise specifications to assist the
synthesis process.

@@

Expand on earlier points. Start from a functional language with linear
types ($\tensor$, $\lolli$, $\oplus$, $\bang$); build on it with more
``stuff''. General techniques, drawn from proof theory via props as
types: explore focusing to tame non-determinism/search space. Partial
synthesis is still proof search! Add other techniques as we go along
(e.g. for recursion we need to constrain recursive calls).

Mention that you've already implemented a type-checker for this
(useful as a prelim. Exercise but also later, for \emph{validation}).

Validation and evaluation: validation is as simple as ``does it typecheck''? can you synth? how fast?

\section{Bonus}

\[
  \begin{array}{lcl}
    A, B & ::= & A \tensor B \mid A \lolli B \mid A \with B \mid A
                 \oplus B \mid \bang A \mid \one \mid \dots\\[1ex]
    M,N & ::= & \lambda x{:}A.M\\
         & \mid & M\,N\\
         & \mid & (M \tensor N)\\
         & \mid & \llet{x\tensor y = M}{N} \\
         & \mid & x \\
         & \mid & \dots\\
    \end{array}
\]

\[
  \infer[$(\lolli\! I)$]
  {\Delta , x{:}A \vdash M : B}
  {\Delta \vdash \lambda x {:} A . M : A \lolli B}
  \quad
  \infer[$(\lolli\! E)$]
  {\Delta_1 \vdash M : A \lolli B \and \Delta_2 \vdash N : A}
  {\Delta_1, \Delta_2 \vdash M\,N : B}
\]

\[
  \infer*[left=($\tensor I$)]
  {\Delta_1 \vdash M : A \and \Delta_2 \vdash N : B}
  {\Delta_1 , \Delta_2 \vdash (M \tensor N) : A \tensor B}
  \quad
  \infer*[right=($\tensor E$)]
  {\Delta_1 \vdash M : A \tensor B \and \Delta_2 , x{:}A, y{:}B\vdash
    N : C }
  {\Delta_1 , \Delta_2\vdash \llet{x\tensor y = M}{N} : C }
\]

>>>>

% \section{Separate}

% More about the concepts than citing concrete works (but citations are
% needed!):
% \begin{enumerate}
% \item PL/Type systems in a ``formal'' sense (inference rules, etc) [DONE]
% \item Propositions as types: Proofs as programs, proof search as
%   synthesis, etc. [MOSTLY DONE]
% \item Linear logic: a ``resource aware logic'' [ONGOING]
% \item Proof search in LL: challenges, focusing as a ``solution'' [ONGOING]
% \item Linear $\lambda$-calculus: what does it look like.
% \item Type-based synthesis: ``inverting'' a type system.
% \end{enumerate}

@@

% Refinement types provide very precise specifications to assist the
% synthesis process.

@@

We start with the definition of our type system and simply-typed functional language
with linear types ($\lolli$, $\tensor$, $1$, $\with$, $\oplus$, $\bang$, $Bool?$).
Through the Curry-Howard correspondence, we'll formulate the synthesis of a
linear type in our functional lanugage as proof search in linear logic.
Since \emph{natural deduction} is ``natural'' in practice, and thus, harder to streamline and reason about,
we'll use the equivalent \emph{sequent calculus} to formulate our inference rules.
In sequent calculus, all our rules can be read in a \emph{bottom-up} manner, simplifying the reasoning for the synthetizer.
Therafter, a way to reason about our resource contexts is needed, and the Hodas and Miller's? input-output model~\cite{} comes into play.
According to it, instead of a single context that we must split into parts when proving the judgments
above the inference rule line, we'll utilize an \emph{input} and an \emph{output} context -- to prove $A\tensor B$ with $\Delta$, prove $A$ with $\Delta$, consuming the needed resources, and output a $\Delta'$, prove $B$ with $\Delta'$ and the output the remaining resources in $\Delta''$, and finally, after having proven $A\tensor B$, output $\Delta''$, which might be used by another judgment.
With the input/output contexts, and the sequent calculus formulation of the inference rules, we're on a good path to synthetize programs.
However, even though proof-search here might seem direct, it still encompasses a lot of non-determinism, e.g., which rule to apply at a certain point?
To this we add focusing~\cite{} -- a method created to trim down the search space of valid proofs in a linear logic, by conducting the
proof-search first with invertible rules, and then by focusing where no more direct (invertible) rules can be applied.
With focusing we tame some of non-determinism from the synthesis aspect. \todo{talvez algo mais sobre focusing aqui.}
We might, later on, add refinement types also as a way to constrain the valid programs space.
Additionally, we might need to discuss strategies to synthetize recursive functions as well.

Putting together everything above, and the more detailed information in the background section -- linear types, sequent calculus, resource management, and focusing,
we have a set of inference rules (written out in the appendix) that define how the synthesis process will progress.

% Expand on earlier points. Start from a functional language with linear
% types ($\tensor$, $\lolli$, $\oplus$, $\bang$); build on it with more
% ``stuff''. General techniques, drawn from proof theory via props as
% types: explore focusing to tame non-determinism/search space. Partial
% synthesis is still proof search! Add other techniques as we go along
% (e.g. for recursion we need to constrain recursive calls).

% Mention that you've already implemented a type-checker for this
% (useful as a prelim. Exercise but also later, for \emph{validation}).

% Validation and evaluation: validation is as simple as ``does it typecheck''? can you synth? how fast?

% \section{Bonus}

% \[
%   \begin{array}{lcl}
%     A, B & ::= & A \tensor B \mid A \lolli B \mid A \with B \mid A
%                  \oplus B \mid \bang A \mid \one \mid \dots\\[1ex]
%     M,N & ::= & \lambda x{:}A.M\\
%          & \mid & M\,N\\
%          & \mid & (M \tensor N)\\
%          & \mid & \llet{x\tensor y = M}{N} \\
%          & \mid & x \\
%          & \mid & \dots\\
%     \end{array}
% \]

% \[
%   \infer[$(\lolli\! I)$]
%   {\Delta , x{:}A \vdash M : B}
%   {\Delta \vdash \lambda x {:} A . M : A \lolli B}
%   \quad
%   \infer[$(\lolli\! E)$]
%   {\Delta_1 \vdash M : A \lolli B \and \Delta_2 \vdash N : A}
%   {\Delta_1, \Delta_2 \vdash M\,N : B}
% \]

% \[
%   \infer*[left=($\tensor I$)]
%   {\Delta_1 \vdash M : A \and \Delta_2 \vdash N : B}
%   {\Delta_1 , \Delta_2 \vdash (M \tensor N) : A \tensor B}
%   \quad
%   \infer*[right=($\tensor E$)]
%   {\Delta_1 \vdash M : A \tensor B \and \Delta_2 , x{:}A, y{:}B\vdash
%     N : C }
%   {\Delta_1 , \Delta_2\vdash \llet{x\tensor y = M}{N} : C }
% \]
